Elaboration with First-Class Implicit Function Types

### 2020-07-01 paper.tex
typo
camera version

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous,prologue,dvipsnames]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

@@

%% Title information
\title{Elaboration with First-Class Implicit Function Types}

@@

\subsection{Surface syntax}
Figure \ref{fig:surface} shows the possible constructs in the surface
language. We only have terms, as we have Russell-style universe in the core, and
we can conflate types and terms for convenience. The surface syntax does not
have semantics or any well-formedness relations attached; its sole purpose is to
serve as input to elaboration. Hence, the surface syntax can be also viewed as a
small untyped tactic language which is interpreted by the elaborator.

@@

\subsection{Core syntax}
\label{sec:core}

@@

\begin{definition}[Closing types]
We use curried function types to close over record types in the scope. If a
bound variable does not have a record type, then we do as before\footnote{This
  implies that we close over meta-headed types using plain functions. In theory,
  this causes a higher-order version of the basic implicit insertion problem: we
  are uncertain about whether we should be uncertain about implicit
  insertions. This issue does not seem to come up in realistic programs in the prototype implementation; the first-order treatment of uncertain insertions seem to leave little remaining ``higher-order'' uncertainty on the table }. We prepend the
following clause to Definition
\ref{def:closingtype}:
\[
  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
\]
\end{definition}
\begin{definition}[Contextualization]
We extend spine notation to applications of curried functions. For example, we
may have a spine $\ol{t} \equiv (\{x : \ol{A}\},\,\{y : \ol{B}\})$. We
accordingly revise Definition \ref{def:contextualization} for
$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
for each record type in $\Gamma$.
\end{definition}

@@

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
&\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow
            \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}) \\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term and wrap the result in a strictly curried
$\lambda$. We again need to create new constancy constraint. This way, if $B$
does not depend on $x : \Rec\,A$, the strictly curried $\lambda$ in the output
immediately computes away, since $A$ is solved to $\epsilon$.

@@

\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}}{(\mi{List}\,(\{A : \U\}\to A \to A))}
      \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}} \\
  & \scriptstyle{2 }  && \qquad \einfert{\mi{cons}\,(\lambda\,x.\,x)} \\
  & \scriptstyle{3 }  && \qquad\quad \einfert{\mi{cons}}\\
  & \scriptstyle{4 }  && \qquad\quad = \mi{cons}\,\{\alpha_0\}\,:\,\alpha_0 \to \mi{List}\,\alpha_0 \to \mi{List}\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = \mi{cons}\,
                       \{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,
                       (\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\\
  & \scriptstyle{8 }  && \qquad \echeckt{\mi{nil}}{(\mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))}\\
  & \scriptstyle{9 }  && \qquad = \mi{nil}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\\
                       & \scriptstyle{10} && \quad = \mi{cons}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,(\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\,\\
                       & && \hspace{4.4em} (\mi{nil}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\})\\
  & && \hspace{3em} : \mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{11} && \quad \unify\,(\mi{List}\,(\{A : \U\}\to A \to A))\,(\mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{13} && \qquad\quad \unify\,\alpha_1\,((A : \U) \TCons (\alpha_3\,A))\\
  & \scriptstyle{14} && \qquad \quad \unify\,(A \to A)\,(\{z : \ol{\alpha_3\,A}\} \to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}\to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\})\\
  & \scriptstyle{15} && \qquad \qquad \unify\,(\alpha_3\,A)\,\epsilon\\
  & \scriptstyle{16} && \qquad \qquad \unify\,(A \to A)\,(\alpha_2\,\{A\} \to \alpha_2\,\{A\})\\
  & \scriptstyle{17} && \qquad \qquad \quad \unify\,A\,(\alpha_2\,\{A\})\\
  & \scriptstyle{18} && \qquad \qquad \quad \unify\,A\, A\\
  & \scriptstyle{19} && = \mi{cons}\,\{\{A : \U\}\to A \to A\}\,(\lambda\,\{A\}\,x.\,x)\,(\mi{nil}\,\{\{A : \U\}\to A \to A\})
\end{alignat*}
We diverge from the previous attempt at line $\scriptstyle{5}$. Here, we check
$\lambda\,x.\,x$ with the meta $\alpha_0$, so we create a fresh $\alpha_1 :
\Tel$ meta and wrap the result as $\lambda\,\{y :
\ol{\alpha_1}\}.\,\lambda\,x.\,x$. This result has the inferred type $\{y :
\ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y :
\ol{\alpha_1}\}$, and we promptly unify $\alpha_0$ (which stands for the list
element type) with it. On line $\scriptstyle{10}$, we return the elaborated
$\mi{cons}$ expression, where the list element type is made explicit. It only remains
to unify the inferred and expected types. On line $\scriptstyle{12}$ we have the
case where a curried function type is matched with an implicit function type, so
on line $\scriptstyle{13}$ we refine $\alpha_1$ to a non-empty telescope, and
proceed unifying the domains. Note that on line $\scriptstyle{14}$ we have
$\alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}$, which results from applying the
substitution $y \mapsto (A :: z)$, and computing $\alpha_2\,\{(A :: z) : \ol{(A :
  \U) \TCons (\alpha_3\,A)}\}$ further using the \LabTirName{app-$\TCons$} rule
from Figure \ref{fig:telescopes}.

@@

\begin{acks}
  This work was supported by the European Union, co-financed by the
  European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002) and COST Action
  EUTypes CA15123.
\end{acks}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,screen,dvipsnames]{acmart}\settopmatter{}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% %% Journal information
%% %% Supplied to authors by publisher for camera-ready submission;
%% %% use defaults for review submission.
%% \acmJournal{PACMPL}
%% \acmVolume{1}
%% \acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
%% \acmArticle{1}
%% \acmYear{2018}
%% \acmMonth{1}
%% \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}

\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3408983}
\acmYear{2020}
\copyrightyear{2020}
\acmSubmissionID{icfp20main-p21-p}
\acmJournal{PACMPL}
\acmVolume{4}
\acmNumber{ICFP}
\acmArticle{101}
\acmMonth{8}
\startPage{1}

%% %% Copyright information
%% %% Supplied to authors (based on authors' rights management selection;
%% %% see authors.acm.org) by publisher for camera-ready submission;
%% %% use 'none' for review submission.
%% \setcopyright{none}
%% %\setcopyright{acmcopyright}
%% %\setcopyright{acmlicensed}
%% %\setcopyright{rightsretained}
%% %\copyrightyear{2018}           %% If different from \acmYear

@@

%% Title information
\title{Elaboration with First-Class Implicit Function Types}
\titlenote{The author was supported by the European Union,
co-financed by the European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002).}

@@

\subsection{Surface Syntax}
Figure \ref{fig:surface} shows the possible constructs in the surface
language. We only have terms, as we have Russell-style universe in the core, and
we can conflate types and terms for convenience. The surface syntax does not
have semantics or any well-formedness relations attached; its sole purpose is to
serve as input to elaboration. Hence, the surface syntax can be also viewed as a
small untyped tactic language which is interpreted by the elaborator.

@@

\subsection{Core Syntax}
\label{sec:core}

@@

\begin{definition}[Closing types]
We use curried function types to close over record types in the scope. If a
bound variable does not have a record type, then we do as before\footnote{This
  implies that we close over meta-headed types using plain functions. In theory,
  this causes a higher-order version of the basic implicit insertion problem: we
  are uncertain about whether we should be uncertain about implicit
  insertions. This issue does not seem to come up in realistic programs in the prototype implementation; the first-order treatment of uncertain insertions seem to leave little remaining ``higher-order'' uncertainty on the table.}. We prepend the
following clause to Definition
\ref{def:closingtype}:
\[
  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
\]
\end{definition}
\begin{definition}[Contextualization]
We extend spine notation to applications of curried functions. For example, we
may have a spine $\ol{t} \equiv (\{x : \ol{A}\},\,\{y : \ol{B}\})$. We
accordingly revise Definition \ref{def:contextualization} for
$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
for each record type in $\Gamma$.
\end{definition}

@@

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
&\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow
            \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}) \\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term and wrap the result in a strictly curried
$\lambda$. We again need to create a new constancy constraint. This way, if $B$
does not depend on $x : \Rec\,A$, the strictly curried $\lambda$ in the output
immediately computes away, since $A$ is solved to $\epsilon$.

@@

\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & \scriptstyle{0 }\qquad\qquad && \echeckt{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}}{(\mi{List}\,(\{A : \U\}\to A \to A))}
      \hspace{14em}\\
  & \scriptstyle{1 }  && \quad \einfert{\mi{cons}\,(\lambda\,x.\,x)\,\mi{nil}} \\
  & \scriptstyle{2 }  && \qquad \einfert{\mi{cons}\,(\lambda\,x.\,x)} \\
  & \scriptstyle{3 }  && \qquad\quad \einfert{\mi{cons}}\\
  & \scriptstyle{4 }  && \qquad\quad = \mi{cons}\,\{\alpha_0\}\,:\,\alpha_0 \to \mi{List}\,\alpha_0 \to \mi{List}\,\alpha_0\\
  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
  & \scriptstyle{6 }  && \qquad\quad = \lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x\\
  & \scriptstyle{7 }  && \qquad = \mi{cons}\,
                       \{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,
                       (\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\\
  & \scriptstyle{8 }  && \qquad \echeckt{\mi{nil}}{(\mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))}\\
  & \scriptstyle{9 }  && \qquad = \mi{nil}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\\
                       & \scriptstyle{10} && \quad = \mi{cons}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,(\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\,\\
                       & && \hspace{4.4em} (\mi{nil}\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\})\\
  & && \hspace{3em} : \mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{11} && \quad \unify\,(\mi{List}\,(\{A : \U\}\to A \to A))\,(\mi{List}\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))\\
  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
  & \scriptstyle{13} && \qquad\quad \unify\,\alpha_1\,((A : \U) \TCons (\alpha_3\,A))\\
  & \scriptstyle{14} && \qquad \quad \unify\,(A \to A)\,(\{z : \ol{\alpha_3\,A}\} \to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}\to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\})\\
  & \scriptstyle{15} && \qquad \qquad \unify\,(\alpha_3\,A)\,\epsilon\\
  & \scriptstyle{16} && \qquad \qquad \unify\,(A \to A)\,(\alpha_2\,\{A\} \to \alpha_2\,\{A\})\\
  & \scriptstyle{17} && \qquad \qquad \quad \unify\,A\,(\alpha_2\,\{A\})\\
  & \scriptstyle{18} && \qquad \qquad \quad \unify\,A\, A\\
  & \scriptstyle{19} && = \mi{cons}\,\{\{A : \U\}\to A \to A\}\,(\lambda\,\{A\}\,x.\,x)\,(\mi{nil}\,\{\{A : \U\}\to A \to A\})
\end{alignat*}
\endgroup

We diverge from the previous attempt at line $\scriptstyle{5}$. Here, we check
$\lambda\,x.\,x$ with the meta $\alpha_0$, so we create a fresh $\alpha_1 :
\Tel$ meta and wrap the result as $\lambda\,\{y :
\ol{\alpha_1}\}.\,\lambda\,x.\,x$. This result has the inferred type $\{y :
\ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y :
\ol{\alpha_1}\}$, and we promptly unify $\alpha_0$ (which stands for the list
element type) with it. On line $\scriptstyle{10}$, we return the elaborated
$\mi{cons}$ expression, where the list element type is made explicit. It only remains
to unify the inferred and expected types. On line $\scriptstyle{12}$ we have the
case where a curried function type is matched with an implicit function type, so
on line $\scriptstyle{13}$ we refine $\alpha_1$ to a non-empty telescope, and
proceed unifying the domains. Note that on line $\scriptstyle{14}$ we have
$\alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}$, which results from applying the
substitution $y \mapsto (A :: z)$, and computing $\alpha_2\,\{(A :: z) : \ol{(A :
  \U) \TCons (\alpha_3\,A)}\}$ further using the \LabTirName{app-$\TCons$} rule
from Figure \ref{fig:telescopes}.

@@
